---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# automerge

<!-- badges: start -->
<!-- badges: end -->

R Bindings for the Automerge CRDT Library

## Overview

`automerge` provides R bindings to the [Automerge](https://automerge.org/) Conflict-free Replicated Data Type (CRDT) library via its C FFI. Automerge enables automatic merging of concurrent changes without conflicts, making it ideal for:

- Distributed systems
- Collaborative applications
- Offline-first architectures
- Cross-platform data synchronization

## Features

- Full support for Automerge data types: maps, lists, text, counters
- Intuitive functional API with S3 methods
- Low-level and high-level synchronization protocols
- Seamless interoperability with JavaScript and other Automerge implementations
- Zero runtime dependencies (only base R)

## Installation

### From Source (Development Version)

```r
# Install from GitHub
pak::pak("posit-dev/automerge-r")
```

### System Requirements

To build from source, you need:

- R >= 4.2
- Rust toolchain >= 1.89.0 - Install from https://rustup.rs/
- CMake >= 3.25 - Included in Rtools43+ on Windows

Alternatively, install `automerge-c` system-wide to skip the source build.

## Status

🚧 **Under Active Development** 🚧

This package is currently in the initial development phase. Core functionality is not yet implemented and any existing functionality is subject to change at any time.

## Example Usage

### Creating and Modifying Documents

Use familiar R syntax to work with Automerge documents:

```{r}
library(automerge)

doc <- am_create()
doc$name <- "Alice"
doc$age <- 30L
doc[["active"]] <- TRUE
doc
```

### Reading Values

Access values just like a regular R list:

```{r}
doc$name # "Alice"
doc[["age"]] # 30L
```

### Nested Structures

Create complex nested objects in a single call:

```{r}
doc$user <- list(
  name = "Bob",
  age = 25L,
  address = list(city = "NYC", zip = 10001L)
)
```

Nested objects work independently:

```{r}
user <- doc$user
user$email <- "bob@example.com"
user
```

### Advanced Types

Automerge supports specialized CRDT types:

```{r}
doc$created <- Sys.time() # POSIXct timestamp
doc$score <- am_counter(0) # Counter
doc$notes <- am_text("new") # Text object with CRDT semantics
```

### Working with Text

Text objects support fine-grained editing:

```{r}
text_obj <- doc$notes
am_text_splice(doc, text_obj$obj_id, 3, 0, "s")
am_text_get(doc, text_obj$obj_id)
```

### Utility Methods

Standard R operations work as expected:

```{r}
length(doc) # Number of keys
names(doc) # Key names
as.list(doc) # Convert to R list
```

### Saving and Loading

Persist documents to disk or transfer over network:

```{r}
doc |> am_commit("Initial data")
bytes <- am_save(doc)
doc2 <- am_load(bytes)
doc2
```

### Forking and Merging

Create independent copies and merge changes automatically:

```{r}
doc3 <- am_fork(doc)
doc3$name <- "Charlie"

am_merge(doc, doc3)
doc$name
```

### Low-Level API

For fine-grained control, use the functional API:

```{r}
doc <- am_create()
am_put(doc, AM_ROOT, "name", "Alice")
am_put(doc, AM_ROOT, "age", 30L)
doc
```

## Resources

- [Automerge Website](https://automerge.org/)
- [Automerge GitHub](https://github.com/automerge/automerge)
